\chapter{Concurrency and Thread-Safety}

\label{Unicon-Concurrency-Thread_Safety}
A {\em thread\/} is a fundamental building block of an executing
program, consisting of a set of registers (including a program counter
indicating what instruction it is executing) and a stack with which
to evaluate intermediate results, place parameters and perform
function calls, etc.  If you have a running program, you have one
thread; if you have multiple threads, you might have one running at
a time and the threads taking turns, or you might have more than one
running simultaneously, a condition known as concurrent execution.

Icon's co-expression type is a thread-type that does not allow
concurrent execution.  In fact, the current implementation of Icon
uses posix threads with simple rules to ensure that only one of them
at a time can ever execute.

Since the turn of the 21st century, however, concurrency has become
essential to utilizing modern computer processors.  Once CPU
manufacturers hit physical limits and could no longer double the
clock speed every 18 months, they have turned a large part of their
attention to the addition of more and more execution cores at a given
clock speed.  Because multiple cores are now ubiquitous, and
concurrent programming is notoriously difficult, the pressure on
very-high level languages to support concurrency is high, but
adding concurrency to a very high-level language is more difficult
than it might seem.  Most popular mainstream scripting languages
offer pseudo-concurrency, but true concurrency is often limited by
a ``global interpreter lock'', or GIL, that prevents simultaneous
access by multiple threads to the same virtual machine.

In order to provide true concurrency, Unicon extends Icon's posix
threads co-expression implementation, and provides a thread-safe
runtime system.  Some of this was accomplished by switching out unsafe
C library functions for their thread-safe counterparts, but a far more
substantial part of it was accomplished by making threads as
independent of each other as possible, giving each thread its own copy
of various interpreter state variables that formerly were
globals. This is one of the few semantic differences between Icon and
Unicon.

\section{Thread State}

Unicon's threads implementation was influenced heavily by the
earlier development of dynamic loading and the ability to execute
multiple programs as co-expressions within the same virtual machine.
That earlier implementation produced a notion of a ``current program
state'' embodied by a \texttt{struct progstate}. Concurrent threads
share parts of what was the progstate, while giving each thread its
own copy of as many parts as possible in order to minimize threads'
interference with each other.  Those parts of the global state that
are replicated on a per-thread basis form \texttt{struct threadstate}
in \textfn{src/h/rstructs.h}.

From a language design standpoint, adding concurrency to Unicon consisted of:
\begin{enumerate}
\item
  adding a way to tell co-expressions to execute concurrently instead of
  waiting to be activated.
\item
  extending the activation operator \texttt{@} and adding new
  operators to provide a simple mechanism for thread synchronization
  and communication. These were achieved by introducing a new reserved
  word \texttt{thread} and extending the \texttt{@} operator to
  maintain producer-consumer queues for asynchronous operation in
  addition to operators for new forms of synchronization and
  communication as discussed later in this section. The thread syntax
  can be used in the following manner:
\end{enumerate}

\begin{iconcode}
thread expr
\end{iconcode}

The language was also extended with several new built-in functions for
high-level access to mutual exclusion primitives and condition variables,
in addition to extending a few existing functions to support new semantics.
For example, the built-in function \texttt{wait()} was extended to allow a
thread argument to support the semantics of \texttt{join} in thread
programming. Many of the extensions have to do with synchronization. Some
problems require using advanced synchronization mechanisms and rely on the
language support to achieve full control over the execution of threads and
protect shared data. This section covers the synchronization techniques
introduced into the Unicon language. It also covers thread communication
and the new communication operators.

\section{Synchronization}
\subsection{Critical Regions and Mutexes}
A mutex is a synchronization object used to protect shared data and
serialize threads in critical regions. For example when two threads compete
to increment a variable, the end result might not be what the programmer
intended. In such cases a mutex should be used to protect access to the
variable. Any operation or data where more than one thread can execute
non-deterministically, leading to data corruption or incorrect results is
called thread unsafe. Thread unsafe code or data structures have to be
handled correctly via synchronization mechanisms to achieve correct
behavior with correct results.  A mutex object can be created using the
\texttt{mutex()} function. The returned mutex object can then be
locked/unlocked via the functions \texttt{lock()} and \texttt{unlock()} to
serialize execution in a critical region. Unicon provides a special syntax
for critical regions which is equivalent to a
\texttt{lock()}/\texttt{unlock()} pair, that aims mainly to guarantee that
a mutex is released at the end of a critical region, beside enhancing the
readability of the program. Here is the syntax:

\iconline{critical mtx: expr}

\noindent
This is equivalent to:

\begin{iconcode}
lock(mtx)\\
expr\\
unlock(mtx)\\
\end{iconcode}

Mutexes do not exist solely as independent objects, they are also used as
attributes of other objects, namely attributes of the mutable data
types. Any data structure in Unicon that can be used in a thread unsafe
manner is better protected by a mutex. Instead of declaring a separate
mutex, locking and unlocking it, the structure can just be marked as needs
a mutex/protection and the language does an implicit locking/unlocking,
protecting the operations that might affect the integrity of the data
structure.  For example, if several threads are pushing and popping
elements into and out of a list, this creates thread-unsafe operations on
the list that require protection.  Using a thread-safe list results in less
lines of code compared with doing explicit locking, and also produces a
more efficient program doing less locking and unlocking at the language
level, or even not doing at all. For example, a list can be protected by
passing it to the function \texttt{mutex()} as follows:

\iconline{mutex(L)}

\noindent
From this point on, any puts to the list or gets from \texttt{L} are
protected by an implicit mutex locking mechanism. This helps make
concurrent programming almost as easy as writing a sequential program.  The
only needed step is to notify the language at the beginning that the data
structure is shared, by passing it to the \texttt{mutex()} function. Having
to explicitly mark structures as shared ensures that mutexes are only used
where needed, and not wasting time on unnecessary locking/unlocking for
structures that are not shared. The locking/unlocking of protected
structures slows down structures operations by about 7\%. The
\texttt{mutex()} function takes a second optional parameter denoting an
existing mutex object or an object that is already marked as shared (has a
mutex attribute) . Instead of creating a new mutex object for the data
structure, the existing mutex is used as an attribute for the data
structure. If the second object is a structure that is not marked as
shared, a new mutex is created. This is useful when two objects need to be
protected by the same mutex. For example, the list \texttt{L} and the table
\texttt{T} in the following example share the same mutex:

\begin{iconcode}
\>mtx := mutex()\\
\>L := mutex([ ], mtx)\\
\>T := mutex(table(), mtx)\\
\end{iconcode}

\noindent
which is equivalent to the following if the mutex does not need to be explicit:

\begin{iconcode}
\>L := mutex([ ])\\
\>T := mutex(table(0), L)\\
{\textnormal {or}}\\
\>L := [ ]\\
\>T := mutex(table(0), L)\\
\end{iconcode}

In all cases, \texttt{lock(L)} and \texttt{lock(T)} lock the same mutex,
serializing execution on both data structures.  Not all operations on data
structures produce correct results, only atomic operations do. In other
words, implicit locking/unlocking takes place per operation, which means
even if each one of the two operations is safe, the combination might not
be. A critical region is still needed to combine the two.  For example, if
\texttt{L[1]} has the value of 3 and two threads are trying to increment
\texttt{L[1]} as the following:

\iconline{L[1] := L[1] + 1}

\noindent
the result in \texttt{L[1]} could be 4 or 5. That is because reading
\texttt{L[1]} (the right side of the assignment) and storing the result
back in \texttt{L[1]} are two different operations (not atomic) and are
actually separated in time. The good news is that solving such an issue
does not require an extra explicit mutex.  If \texttt{L} is marked as
shared (passed to the \texttt{mutex()} function ) it can be passed to
\texttt{lock()}/\texttt{unlock()} functions.  It can be used with the
critical syntax like the following:

\iconline{critical L: L[1] := L[1] + 1}

\subsection{Initial clause}
\label{Unicon-Initial-Clause}
A procedure in a Unicon (or Icon) program can have an initialization clause
at the top of the procedure.  This gets executed only once the first time
the procedure is entered.  The initial clause provides a very convenient
way to have local static variables and their initialization in the same
procedure instead of relying on global variables and having to initialize
them somewhere else.  An Icon procedure that produces a sequence of numbers one
at each call can be written as:

\begin{iconcode}
procedure seq()\\
\>static i\\
\>initial i:=0\\
\>i := i+1\\
\>return i\\
end\\
\end{iconcode}

In Unicon the \texttt{initial} clause must be thread-safe. It can be
thought of as a built-in critical region that is run only once. No thread
is allowed to enter the procedure if there is a thread currently still
executing in the \texttt{initial} block. This can be useful specifically in
a concurrent environment to do critical initialization such as creating a
new mutex object instead of declaring a mutex variable to be global and
initializing it somewhere else or passing it from one function to another
where it will be actually used. The procedure above is correct for Icon but
in a concurrent Unicon program it is not thread-safe, even with the
\texttt{initial clause}.  A correct concurrent version of \texttt{seq()}
would look like this:

\begin{iconcode}
procedure seq()\\
\>local n\\
\>static i, region\\
\>initial \{i:=0; region:=mutex()\}\\
\>critical region: n := i := i+1\\
\>return n\\
end \\
\end{iconcode}

Note the use of the local variable \texttt{n} to temporarily hold the value
of the counter \texttt{i} while still in the critical region. That is
because once the thread leaves the critical region, there is no guarantee
that the value of \texttt{i} would remain the same before returning the
value of \texttt{i}.  Using the variable \texttt{n} guarantees that the
value returned is correct even if the value of \texttt{i} is changed by
another thread.

\subsection{Condition Variables}
Mutexes protect shared data in critical regions, and block threads if more
than one thread tries to enter the critical region. Condition variables
provide thread blocking/resumption that is not tied to accessing shared
data like a mutex. A condition variable allows a thread to block until an
event happens or a condition changes. For example, in a producer/consumer
problem, a consumer keeps spinning to get values out of the shared list. In
similar situations in real life applications, any spinning could be a waste
of resources; other threads including producer threads could be using the
resources doing something useful instead. The consumer should block until
there is data to process in the list. This is where a condition variable
comes into play.  A condition variable can be created using the function
\texttt{condvar()}.  The return object is a condition variable that can be
used with \texttt{wait()} and \texttt{signal()}
functions. \texttt{wait(cv)} blocks the current thread on the condition
variable \texttt{cv}. The thread remains blocked until another thread does
a \texttt{signal(cv)}, which wakes up one thread blocked on \texttt{cv}. A
very important aspect of using a condition variable is that the condition
variable is always associated with a mutex. More specifically, the
\texttt{wait()} function has to be always protected by a mutex.  Unicon
provides a built-in mutex for condition variables which can be thought of
as an attribute, similar to thread safe data structures. This means a
condition variable can also be used with \texttt{lock()}/\texttt{unlock()}
functions or the critical clause. It is important to realize that not only
\texttt{wait()} has to be protected by a critical region, but also the
condition or the test that leads a thread to wait on a condition
variable. The \texttt{signal()} function takes a second optional parameter
denoting the number of threads to be woken up.  By default, that number is
one, but it can be any positive value. From example:

\iconline{every !4 do signal(cv)}

\noindent
Can be written as:

\iconline{signal(cv, 4)}

\noindent
Furthermore, if all of the threads waiting on \texttt{cv} needs to be woken
up, a special 0 ( or \texttt{CV\_BROADCAST}) value can be passed to
\texttt{signal()} causing it to broadcast a wakeup call for all threads
waiting on \texttt{cv}:

\iconline{signal(cv, 0) }
\noindent or
\iconline{signal(cv, CV\_BROADCAST)}

\section{Thread Communication}
Co-expressions, as explained in \ref{Overview-CoExpressions}
(page \pageref{Overview-CoExpressions}), have a simple form
of communication influenced by their synchronous execution. Threads on the
other hand, run in parallel, creating a dynamic environment for
communication. In many cases, a running thread must send a value to another
thread without waiting for results, or receive a value from another thread
if there is one without waiting. The \texttt{@} operator is not suitable
for this kind of (asynchronous) communication. Threads communication needs
to go well beyond co-expressions, for which a single \texttt{@} operator
provided communication bundled with synchronization. Unicon adds four
operators for asynchronous communication. These are \texttt{@>},
\texttt{@>{}>}, \texttt{<@} and \texttt{<{}<@}. The operators correspond to
send, blocking send, receive and blocking receive.

\subsection{Thread messaging Queues}
Thread communication is done through messaging queues that are initialized at thread
creation. Each thread maintains two queues called the inbox and outbox that are created
with the thread. When a thread sends a message with an explicit destination, the message
is queued in the destination's inbox. Otherwise it is queued into the sender's outbox.
A thread can receive messages from another thread by dequeuing messages from the source's
outbox if there is an explicit source, otherwise it dequeues messages from its own inbox.
Figure \ref{msg-queues} presents two threads with the inboxes and outboxes.

% [DonW] The order in which things are drawn in this picture is important
%        because some parts of an object must be concealed and other parts
%        must overlay other objects. The conventional TikZ way of dealing
%        with this (using a background layer) doesn't work in this situation:
%        we must draw the objects on a single path and use "behind path".
\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[draw,very thick,font=\small\tt,
    % general 3D effect with light coming from the top, and a shadow
    3D/.style={text=white,draw=#1!50!white,
       preaction={drop shadow={shadow scale=1.05,shadow xshift=0.75mm,shadow yshift=-0.75mm}},
       top color=#1!5,bottom color=#1!80!black, minimum width=2.2cm},
    % A Thread box
    Thread/.style={3D=#1,shape=rectangle,rounded corners=1.6ex, minimum size=3cm},
    % A message box with arrow
    MA/.style={3D=#1, shape=arrow box, arrow box arrows={east:6mm},minimum height=5mm},
    % A message
    M/.style={3D=#1, shape=ellipse, text=black, shadow scale=0.1}
]
%  \draw[help lines] (0,0) grid (16,8);
  \draw (1,5) node[MA=cyan!90!black] (InA)  {~~~inbox};

  \node[Thread=red, anchor=north west](A) at ($(InA.east) + (0,5mm) $)  {Thread A};
  % We want an ellipse with an arrow coming out of it. TikZ doesn't have one of those,
  % so we fake it with an ellipse and an arrow box behind it (so only the arrow shows).
  \draw node[M=green!60!yellow, name=c1] at ($ (A) + (0,1) $) {~x~@>~B~}
        node[MA=green!60!yellow,minimum width=1mm,
             behind path,arrow box arrows={east:3cm},name=c2] at (c1) {};
  % Similar fakery to produce an inbox connected to another thread with
  % with an extra black arrow coming out of the inbox
  \node[MA=cyan!90!black,anchor=west,name=InB] at (c2.east) {~~~inbox}
        node[Thread=orange,anchor=north west, behind path](B)
             at ($(InB.east) + (0,5mm) $)  {Thread B}
        node[MA=black,draw=black,
             top color=black,bottom color=black,minimum height=1mm, minimum width=1mm,
             arrow box shaft width=0.4,
             arrow box arrows={east:19.5mm},behind path] at (InB) {};

  % We want an outbox with an extra black arrow
  \draw node[MA=red!40!blue,anchor=west, name=OA]
             at ($ (A.south east) + (0,5mm) $) {~~~~outbox}
        node[MA=black,draw=black,
             top color=black,bottom color=black,minimum height=1mm, minimum width=1mm,
             arrow box shaft width=0.4,
             arrow box arrows={east:35.5mm},behind path] at (OA) {};

  \node[MA=red!40!blue,anchor=west, name= OB]
        at ($ (B.south east) + (0,5mm) $) {~~~outbox};
  \node[M=green!60!yellow,anchor=west,minimum width=1mm] at (c2.east) {x};
  \draw node[M=green!60!yellow, name=c3] at ($ (A) - (0,1) $) {~y~@>~}
        node[MA=green!60!yellow,minimum height=1mm, minimum width=1mm,
             arrow box arrows={east:13mm},
             behind path,name=c4] at (c3) {};

  \node[M=green!60!yellow,anchor=west,minimum width=1mm] at (c4.east) {y};
  \node[M=white!80!black,name=c5] at ($ (B) + (0,1) $) {~~~<@~~~};
  \node[M=white!80!black,name=c6] at ($ (B) - (0,1) $) {~~~<@A~~};
\end{tikzpicture}
\end{center}
\caption{\label{msg-queues}Threads messaging queues}
\end{figure}

\subsection{Send and Receive Operators}
The first two operators are \texttt{@>} (send) and \texttt{<@}
(receive). These operators communicate send/receive messages between
threads. The semantics support co-expressions as well which also have
inboxes and outboxes. It does not matter whether threads or co-expressions
are involved in the communication, the semantics are the same. The send
operator has the syntax

\iconline{x@>T}

\noindent
\texttt{x} can be any data type including null, which is equivalent to
omitting it. \texttt{T} refers to a thread to which \texttt{x} is
transmitted. \texttt{x} is placed in \texttt{T}'s inbox. \texttt{x} can be
picked by \texttt{T} using the receive operator which is presented
later. The send operator can also have no destination such as:

\iconline{x@>}

\noindent
In this case \texttt{x} is sent to no one, instead it is placed in the
sender's outbox.  The operator can be read as ``produce \texttt{x}''.
\texttt{x} can then be picked up later by any thread wanting a value from
this sender. For example the sender thread in this case might be a thread
creating prime numbers and placing them in its outbox to be ready for other
worker threads.

The receive operator is symmetric to the send operator and takes two forms, with
explicit source or with no source as follows:

\begin{iconcode}
<@T\\
<@\\
\end{iconcode}

The first case reads: receive a value from \texttt{T}, which gets a value
from \texttt{T}'s outbox, a value produced by \texttt{T}. Going back to the
prime number example mentioned above, \texttt{<@T} would be the way to pick
a prime number produced by the prime number generator thread
\texttt{T}.\texttt{ <@} on the other hand reads values directly from the
receiver's inbox.  It reads messages sent directly to the current thread
doing the receive.  Both \texttt{@>} and \texttt{<@} can succeed or
fail. In the case of \texttt{<@} the operator succeeds and returns a value
from the corresponding queue (inbox/outbox) depending on the operand if the
queue is not empty. If the queue is empty the operation fails directly.  In
the case of \texttt{@>}, if the value is placed in the corresponding queue,
the operation succeeds and returns the size of the queue. If the queue is
full, the send operation fails. The inbox/outbox for each thread is
initialized to have a limited size (it can hold up to 1024 values by
default). This limit can be increased or decreased depending on the
application needs. The limits are useful so that queue sizes do not explode
quickly by default and also to have an implicit
communication/synchronization as explained later in this section.

Each thread has exactly one inbox and one outbox, and each operator call is
mapped to only one of these inboxes or outboxes as seen in Figure \ref{msg-queues}.
All messages from all threads coming to thread B in the figure
end up in its inbox. All threads trying to receive messages from A compete
on its outbox. Both the inbox and the outbox queues are public
communications channels, and it is impossible to distinguish the source of
a message if there are several threads sending messages to the same thread
at the same time. Furthermore, if \texttt{<@} has an explicit source like A in
Figure \ref{msg-queues} it only looks in A's outbox, and does not see messages from A
coming directly to the inbox. Applications that require the sender's
address can attach that information to messages by building them as records
with two fields, one field for data and the other containing the sender's
address. A better approach for private communications for some applications
is the use of lists shared between the two communicating threads or the use
of private communication channels discussed later in this chapter.

\subsection{Inbox/Outbox and the Attrib() Function}

As seen in previous subsections, communications between threads is done through
inbox/outbox queues which are governed by size limits. The size limit (defaults to 1024)
and the actual size dictate how synchronization is melted with the communication.
The size operator \texttt{*} can be used with a thread to query its actual outbox size
(how many values it contains, not the maximum limit) and can be used as follows:

\iconline{outbox\_size := *T}

\noindent
But this is only a single attribute for one queue. To access or change the
different attributes, a new function was introduced to Unicon:
\texttt{Attrib()}, which uses the form:

\iconline{Attrib(handle, attribcode, value, attribcode, value, ...)}

\noindent
The integer codes used by the \texttt{Attrib()} function are supplied by an
include file (\textfn{threadh.icn}) with \texttt{\$define} symbols for the
attributes. This header file is part of a new thread package that is part
of the Unicon distribution called package threads.  It can be imported to a
program via:

\iconline{import threads}

\noindent
When values are omitted, \texttt{Attrib()} generally returns attribute values.
To get the size of the outbox (similar to what the \texttt{*} operation does in the example
presented earlier), the code is:

\iconline{outbox\_size := Attrib(T, OUTBOX\_SIZE)}

\noindent similarly,

\iconline{inbox\_size := Attrib(T, INBOX\_SIZE)}

\noindent gets the current size  of the inbox. On the other hand,

\iconline{Attrib(T, INBOX\_LIMIT, 64, OUTBOX\_LIMIT, 32)}

\noindent sets the inbox and outbox size limits to 64 and 32
respectively. Table \ref{Q-attributes} summarizes all of the available attributes and
their meanings.

\begin{table}[h]
\begin{tabular}{>{\texttt\bgroup}l<{\egroup}ll}
{{\textnormal{Attribute}}} & Meaning & Read/Write\\
INBOX\_SIZE   & Number of items in the inbox & Read Only\\
OUTBOX\_SIZE  & Number of items in the outbox& Read Only\\
INBOX\_LIMIT  & The maximum number of items allowed in the inbox & Read/Write\\
OUTBOX\_LIMIT & The maximum number of items allowed in the outbox & Read/Write\\
\end{tabular}
\caption{\label{Q-attributes}Thread's communication queues attributes}
\end{table}

\subsection{Blocking Send and Receive}
In many situations, senders and receivers generate and consume messages at
different speeds. For example if there is thread that generates prime
numbers very fast to be used by other slow threads, the prime number
generator thread should not continuously generate prime numbers, but create
a batch and wait until it is all used before generating another
batch. Instead of overloading slow receivers or busy waiting for slow
senders, the two ends of communication require a synchronizing mechanism to
tell when to send new messages or when a new message is available. Two more
send and receive operators provide such functionality, these are the
blocking send operator \texttt{@>{}>} and the blocking receive operator
\texttt{<{}<@}. These operators can be used in the same way\texttt{ @>}
and\texttt{ <@} are used, except that instead of failing when the operation
cannot be completed, the new operators block and wait until the operation
succeeds. This is useful in situations where a value is required to
proceed, allowing the thread to block instead of spinning waiting for a
value.

In some cases, a thread might want to use a blocking receive to get values from
a second thread, but it is not willing to block indefinitely; probably the thread
can do some other useful work instead of waiting. The \texttt{<{}<@}  accepts a timeout parameter
to impose a limit on how long to wait for a result before giving up. Here is how
\texttt{<{}<@ }would look like in this case:

\iconline{result := timeout <{}<@\ \ \ \ \#  get from the current thread inbox}
\noindent or
\iconline{result := timeout <{}<@ T\ \ \ \# get from T's outbox}

The timeout operand is a non-negative integer denoting the maximum time to
wait in milliseconds. Negative integers are treated as a null value, an
indefinite blocking receive. A 0 operand indicates no waiting time, making
the end behavior effectively similar to a non-blocking receive. Table \ref{comsOpTab}
summarizes the different forms of the send and receive operators and their
operands:

\begin{table}[h]
\begin{tabular}{|cc@{\hspace*{5mm}}l|}
  \hline
  Operator & Operands & \multicolumn{1}{c|}{Behavior}\\
  \hline
  \multirow{2}{*}{\shortstack{\texttt{@>}\\(send)}}
      & \texttt{msg@>} &
      Place \texttt{msg} in the current thread's outbox, fail if the outbox is full\\
      \cline{2-3}
      & \texttt{msg@>T}
      & Place \texttt{msg} in \texttt{T}'s inbox, fail if \texttt{T}'s inbox is full\\
  \hline
  \multirow{2}{*}{\shortstack{<@\\(receive)}}
      & \texttt{<@}
      & Get a message from the current thread's inbox, fail if it is empty\\
      \cline{2-3}
      & \texttt{<@T}
      & Get a message from \texttt{T}'s outbox, fail if \texttt{T}'s outbox is empty\\
  \hline
  \multirow{2}{*}{\shortstack{@>{}>\\(blocking send)}}
      & \texttt{msg@>{}>}
      & Place msg in the current thread's outbox, block if the outbox is full\\
      \cline{2-3}
      & \texttt{msg@>{}>T}
      &Place msg in \texttt{T}'s inbox, block if the \texttt{T}'s inbox is full\\
  \hline
  \multirow{4}{*}{\shortstack{<{}<@\\(blocking\\receive)}}
      & \texttt{<{}<@}
      & Get a message from the current thread's inbox, block if it is empty\\
      \cline{2-3}
      & \texttt{<{}<@T}
      & Get a message from \texttt{T}'s outbox, block if \texttt{T}'s outbox is empty\\
      \cline{2-3}
      & \texttt{n{<}<@}
      & \multicolumn{1}{p{100mm}|}{
        Get a message from the current thread's inbox, block up to n milliseconds
        if the inbox is empty waiting for new messages}\\
      \cline{2-3}
      & \texttt{n<{}<@T}
      & \multicolumn{1}{p{110mm}|}{
        Get a message from \texttt{T}'s outbox, block up to n milliseconds
        if \texttt{T}'s outbox is empty, waiting for new messages to become available}\\
   \hline
\end{tabular}
\caption{\label{comsOpTab} Summary of the new communication operators}
\end{table}

\subsection{Private Communication Channels}
Inbox and outbox communication queues are visible by all threads all the
time. In some scenarios two or more threads need to communicate with each
other without worrying about other threads sending and receiving messages
using the same shared queues. While it is possible to build a protocol at
the application level on top of the inbox and outbox queues to achieve such
behavior, it is simpler and more efficient to have the threads communicate
privately.  This kind of communication can be done by sharing a list
between two threads and protecting it by an explicit mutex, or using a
thread-safe list. A more formal way for such communication is to use the
\texttt{channel()} function.

Starting a private communication is similar to a network connection, except
that this connection is taking place between two threads in the same
process instead of two different processes that may be on different
machines.  A private communication channel between two threads and can be
created using the function \texttt{channel()}. \texttt{channel()} is part
of the \texttt{threads} package (which was added to Unicon's class
library) and takes one parameter, which is the thread with which the
connection will be initiated.  If \texttt{channel()} succeeds, it returns a
list representing a communication channel between the two threads.
Representing a bidirectional channel that can be used by the two threads,
given that each thread calls the function \texttt{channel()} with the other
thread as an argument. Here is an example.

\noindent
In thread A:
\begin{iconcode}
  chB:= channel(B) | stop("failed to open a channel with B")
\end{iconcode}
In thread B:
\begin{iconcode}
  chA:= channel(A) | stop("failed to open a channel with A")
\end{iconcode}

\noindent
This channel is a {\em directional} communication channel.  One thread
should use it as an outbox, and the other should use it as an inbox.  That
means only one thread will send messages over the channel while the other
receives them from the other end. For many problems, one thread produces
results while one or more threads consume these results, making the
directional nature of the channel suitable for such problems.  In cases
where the communication is needed in both directions, the two communicating
threads can open two channels to send messages in both directions. The
provided channels can be used with the communication operators (all four of
them) with the same semantics as before.  The only difference in this case
is that the right operand is a communication channel instead of a
thread. The \texttt{channel()} function also accepts a second parameter as
an identifier for the channel so that the two ends of the channel find each
other easily without depending on the order in which the \texttt{channel()}
function is called.

\section{Virtual Machine Changes}
The Unicon virtual machine is the Icon virtual machine, extended to better
accommodate object-oriented programming. The extensions made to support
true concurrency would apply equally to the Icon implementation. Other
similar languages face analogous situations, where the lessons learned here
may be applicable. Two main issues had to be addressed in order to allow
concurrent execution of multiple threads in the VM. First, the VM's state
had to be replicated per thread, and second, self-modifying instructions
and the \texttt{initial} clause had to be protected against race
conditions. The two issues are discussed in the following two subsections.

\subsection{VM Registers}
Thread-local storage is utilized to provide each thread with its own copy
of the virtual machine registers, which were formerly global variables in
the VM C code. These include the program counter, stack pointer, the
several frame pointers necessary to manage calls, returns, suspension and
resumption on the stack, and other pieces of virtual machine state.  To add
concurrency, these elements of state are moved into a \texttt{struct
  threadstate} that is allocated for each thread, and keeps all
thread-specific VM state information.  The \texttt{threadstate} is fairly
large, under the assumption that memory is cheap whereas thread
synchronization is expensive. POSIX threads API calls are used to obtain
references to the \texttt{threadstate} structure where it is needed.  A key
concern when using thread-local storage is performance. Initially, all
references to VM registers were replaced, via macros, by references through
a global threadstate pointer variable declared with the compiler storage
specifier \texttt{\_\_thread} that is available in some compilers such as
gcc and Sun's (now Oracle's) cc. This implementation is straightforward but
incurs a substantial performance cost. In addition, the \texttt{\_\_thread}
keyword is not currently available in the \texttt{pthreads} implementation
on OS X and that of Mingw gcc on Microsoft Windows.  Switching to the more
portable pthreads API for thread-local storage using
\texttt{pthread\_getspecific()} allows a faster implementation than
\texttt{\_\_thread}. The straightforward implementation invoked
\texttt{pthread\_getspecific()} once at the top of each C function
containing references to VM register and state variables. The many calls to
this API wherever the VM state is referenced in the runtime system entail
an insignificant performance cost.

\subsection{Self-Modifying Instructions}
The Unicon virtual machine has seven instructions that contain integer
offsets as operands.  These offsets refer to memory addresses within the
instruction region, or to static data regions of the bytecode. They are
used for several types of literals, globals, static variables, and also for
instructions that jump to addresses, such as \texttt{Op\_Goto}. For
performance reasons, when the opcode is first encountered, offsets are
converted to absolute pointers.  The opcode is modified to indicate that
the operand is now a pointer, and the instruction proceeds at full speed on
subsequent executions.

This implementation introduces a race condition when multiple threads
execute the self-modifying instruction at the same time. The problem was
solved by adding a mutex for each self-modifying instruction
opcode. Figure \ref{selfMod-mutex} shows an example self-modifying
instruction code protected with a mutex. Mutex contention could be reduced
further by allocating a separate mutex for each instance of each opcode
(requiring a number of mutexes proportional to the size of the program)
instead of using just seven mutexes for the seven self-modifying opcodes;
it is unlikely that this would be worth implementing, since the number of
instances is large and each instance of a self-modifying instruction uses
its mutex only once when it replaces itself.  To provide a thread safe
\texttt{initial} clause (discussed in \ref{Unicon-Initial-Clause}) at the language
level, a similar technique used for self-modifying instructions was
utilized, unlike self-modifying instructions, where locking and unlocking
take place at the same instructions, with initial clause, the locking
happens at the beginning of the clause, and the unlocking happens at the
end. To allow such behavior, a new VM instruction was introduced which
takes care of updating the instruction at the beginning of the clause, and
unlocks the mutex. The updated instruction causes all of the subsequent
calls to the function containing the initial clause made by any thread to
skip over the clause.  This mechanism effectively eliminates the need to do
any locking/unlocking after it is initialized.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[font=\small\tt]
  %\draw[help lines] (0,0) grid (16,12);
  \def\vsp{0.5cm}
  \def\hsp{1cm}
  %
  % draw the program text
  \begin{scope}[anchor=west]
    \draw
    (0,11)        node[name=L1] {case Op\_Real: ~~~~~~~~~/* real, offset address */}
    ++(0.5,-\vsp) node[name=L2]   {MUTEX\_LOCKID(MTX\_OP\_AREAL);}
    ++(0,-\vsp)   node {}
    ++(0,-\vsp)   node[name=L3]   {if (ipc.op[-1] == Op\_Areal) \{}
    ++(.5,-\vsp)  node[name=L4]      {MUTEX\_UNLOCKID(MTX\_OP\_AREAL);}
    ++(0,-\vsp)   node[name=L5]      {goto L\_areal;}
    ++(-0.5,-\vsp)node[name=L6]   {\}}
    ++(0,-\vsp)   node[name=L7]   {PushVal(D\_Real);}
    ++(0,-\vsp)   node[name=L8]   {opnd = GetWord;}
    ++(0,-\vsp)   node[name=L9]   {opnd += (word)ipc.opnd;}
    ++(0,-0.25)   node {}
    ++(0,-\vsp)   node[name=L10]  {PushAVal(opnd);}
    ++(0,-\vsp)   node[name=L11]  {PutInstr(Op\_Areal, opnd, 1);}
    ++(0,-\vsp)   node[name=L12]  {InterpEVValD((dptr)(rsp-1), e\_literal);}
    ++(0,-\vsp)   node[name=L13]  {MUTEX\_UNLOCKID(MTX\_OP\_AREAL);}
    ++(0,-\vsp)   node[name=L14]  {break;}
    ++(0,-\vsp)   node[name=L15]  {}
    ++(-0.5,-\vsp)node[name=L16] {case Op\_Areal:~~~~~~~~/* real, absolute address */}
    ++(-0.5,-\vsp)node[name=L17]{L\_areal:}
    ++(1,-\vsp)   node[name=L18]  {PushVal(D\_Real);}
    ++(0,-\vsp)   node[name=L19]  {PushAVal(GetWord);}
    ++(0,-\vsp)   node[name=L20]  {InterpEVValD((dptr)(rsp-1), e\_literal);}
    ++(0,-\vsp)   node[name=L21]  {break;};
  \end{scope}
  %
  % annotate the program
  \begin{scope}[every node/.style={single arrow, draw, thin, align=left,
                                   anchor=west,shape border rotate=180},
                color=black!75!white, font=\it ]
    \draw
      (L2.east)+(\hsp,0)     node[text width=6.2cm] {Lock the mutex}
      (L4.east)+(\hsp,0)     node[text width=4.8cm]
        {Check if another thread has done this already}
      (L7.east)+(\hsp,0)     node [text width=9.5cm]
        {Calculate the absolute address using the offset}
      (L11.east)+(3*\hsp,0)  node[text width=5.1cm]
        {Store the absolute address back in the instruction,\break
         replacing the offset}
      (L19.east)+(3*\hsp,0)  node[text width=6.4cm]
        {All of the subsequent references go at full speed;
         no locking or address recalculation is required};
  \end{scope}

\end{tikzpicture}
\end{center}
\caption{\label{selfMod-mutex}Self modifying instruction preceded by a mutex}
\end{figure}

\subsection{Runtime System Support}
Achieving thread-safety in the implementation without a global interpreter
lock required extensive modifications to the runtime system to provide a
thread-safe environment.  This includes safe use of the IO subsystem,
the underlying libraries and, most importantly, heap management in terms of
allocation and garbage collection.

\subsubsection{Input/Output Sub-system}
The Unicon input output sub-system includes very high-level facilities for
accessing files, networks and messaging, 2D and 3D graphics, databases,
pipes and pseudo-ttys. The underlying C library functions implementing
these capabilities are often thread- safe {\color{red} [96]}. However, Unicon
language-level IO operations usually involve several underlying library
calls that must be atomic with respect to threads. Each IO handle is
assigned a mutex when it is opened.  Any IO operation that uses this handle
locks the mutex to guarantee the atomicity of such operations. The cost
incurred by such locks is evaluated in {\color{red} section 7.4}.  The
atomicity of IO is guaranteed only within a single Unicon-level IO
operation. For example \texttt{write(x, y)} is atomic and no other thread
can write to the same file while \texttt{write()} is running.  If multiple
writes or reads are required to be made atomic with respect to other
threads, an explicit lock is needed. In such a case, the IO handle can be
passed to the \texttt{lock()}/\texttt{unlock()} functions to protect it
from any access other than from the current thread during the execution of
several IO operations.

\subsubsection{C Library Thread Safety}
In addition to IO, many C library functions called in the Unicon VM were
implemented before threads were widely used in operating systems. A subset
of those functions is documented in the POSIX standards as
thread-unsafe. About 40 of these function were used in the runtime system.
Most calls to these thread-unsafe functions were replaced with the
thread-safe alternatives available in modern C implementations. A few,
infrequently called, thread-unsafe functions were protected by mutexes. A
couple of the thread-safe alternatives were not portable; new wrapper
functions or implementations were developed in such cases.

\PrimaryIndexBegin{Allocation}
\subsubsection{Allocation and Separate Heaps}
Allocation is a frequent operation for which speed is a top priority in the
Unicon virtual machine.  Unicon programs start with a heap consisting of
one string and one block (structure) region, and allocate more regions as
needed. At any moment, the string and block regions used for allocation are
together referred to as the current heap. If the memory request is bigger
than what is available in the current heap, other regions are checked. If
enough memory is found, the current heap is switched to use the satisfying
region; otherwise a new region is allocated.  With multiple threads, safety
becomes a concern if only one shared heap is available.  When a thread
allocates memory in the current heap and updates the heap's state
variables, the operation must be protected by a mutex. Otherwise different
threads would corrupt each other's allocation requests, leaving the heap in
an invalid state. A similar situation occurs when some operations in the
runtime system deallocate memory after allocating it for a temporary use.

After determining that the locks required in the shared heap strategy were
significantly slowing down the frequent task of memory allocation, the
decision was made to use separate, per-thread heaps.  This allows threads
to use their own private heaps at full speed, just like a single-threaded
application.  Heaps that are not owned by any thread are referred to as
public heaps. Public heaps are a product of threads out-growing their
private heaps. This happens when a thread requests more memory and that
request can only be granted by allocating a new private heap. A program
starts with no public heaps, and with one private heap for the main
thread. A new private heap is allocated for each new thread if none of the
public heap is sufficient. Figure \ref{ThreadHeaps} shows the heap layout
with regards to threads. when a thread finishes, its heap goes back to the
pool of public heaps.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[draw,very thick,font=\small\tt,>=Latex,
    Heap/.style={text=black,draw=red,
      shading=radial, inner color=orange!50!white, outer color=skyblue!70!white,
      shape=rectangle, minimum width=3.5cm, minimum height=9mm},
    Thread/.style={text=black,draw=skyblue,
      shading=radial, inner color=blue!40!white, outer color=orange!50!white,
      shape=rectangle, minimum width=2.5cm, minimum height=8mm, rounded corners=2mm}
  ]
  \definecolor{skyblue}{rgb}{0.53, 0.81, 0.92}
  \definecolor{teagreen}{rgb}{0.82, 0.94, 0.75}
  %\draw[help lines] (0,0) grid (16,10);

  % draw the thread heaps
  \foreach \n/\y in {1/9, 2/7.5, 3/6, n/3}
    {
      \path (3.5,\y) node[rectangle,rounded corners=4mm,fill=teagreen!70!white,
                          minimum width=8cm, minimum height=12mm, name=back-\n] {};
      \path (1,\y) node[Thread,name=Th\n] {Thread \n};
      \path (5,\y) node[Heap,name=Thh\n] {~Thread \n ~heap};
      \draw[red]
        ($(Thh\n.north west)+(0,-1.5mm)$) -- ($(Thh\n.north east)+(0,-1.5mm)$)
        ($(Thh\n.north west)+(5mm,0)$)    -- ($(Thh\n.south west)+(5mm,0)$);
      \draw[->] (Th\n) -- (Thh\n);
    }
    \draw[dotted] ($(back-n.north)+(-2cm,0)$) -- ($(back-3.south)+(-2cm,0)$);
    \node (label) at (3.8,4.5) {Private heaps};
    \draw[->,out=60,in=-120] (label) to (Thh3.south);
    \draw[->,out=-60,in=120] (label) to (Thhn.north);

  % draw the public heaps
  \path (11,6) node[rectangle,rounded corners=8mm,fill=skyblue!30!white,
                          minimum width=4.5cm, minimum height=7.5cm] (P) {};
  \node at ($ (P) + (0,3.2cm) $) {Pool of Public Heaps};

  \foreach \n/\y in {1/2, 2/0.5, m/-2.5}
    {
      \path ($ (P) +(0,\y) $)node[Heap,name=Ph\n] {~Public heap \n};
      \draw[red]
        ($(Ph\n.north west)+(0,-1.5mm)$) -- ($(Ph\n.north east)+(0,-1.5mm) $)
        ($(Ph\n.north west)+(5mm,0)$)    -- ($(Ph\n.south west)+(5mm,0) $);
    }
  \draw[dotted] ($(Ph2.south)+(1,0)$) -- ($(Phm.north)+(1,0)$);

  %connect up the public heaps
  \begin{scope}[<->,arrows={Circle[length=1.5mm,sep]-Latex}]
    \draw[dotted] (Ph2.west)+(1.5mm,0) -- ($(Phm.north west)+(1.5mm,-1mm)$);
    \draw[dotted] (Phm.west)+(3.5mm,0) -- ($(Ph2.south west)+(3.5mm,0)$);
    \draw         (Ph1.west)+(1.5mm,0) -- ($(Ph2.north west)+(1.5mm,-1mm)$);
    \draw         (Ph2.west)+(3.5mm,0) -- ($(Ph1.south west)+(3.5mm,0)$);
  \end{scope}

\end{tikzpicture}
\end{center}
\caption{\label{ThreadHeaps}Threads, private heaps and public heaps}
\end{figure}


All heaps (public and private) are visible to all threads, so any thread can access variables
and data structures in any heap at any time. The accesses can be controlled or protected at the
application level, if needed, in the case of a shared variable. The terms private and public heaps
here are limited to allocating memory in the heaps, and are not to be confused with memory access
to heaps in general. In other words, a thread can allocate memory only in its private heap,
but can access (read and write) variables in all heaps.
Having a full private heap does not mean a thread will automatically allocate a new fresh heap.
When a memory request cannot be granted in the private heap, public heaps are checked to see if
any of them has enough free memory to grant the request. If so, the private heap is swapped with
the public heap, changing the private heap to public and vice-versa. The pool of public heaps is
protected by a mutex. If public heaps do not have enough free memory, a garbage collection will
take place. If the memory request cannot be granted even after that, the thread allocates a fresh heap.
\PrimaryIndexEnd{Allocation}

\PrimaryIndexBegin{Garbage collection}
\subsection{Garbage Collection}

The frequency of garbage collection depends on heap size and application
memory usage patterns.  Historically, many Icon applications ran to
completion without ever garbage collecting. However, modern object oriented
event-driven applications run for longer periods of time, and garbage
collect proportionally often, despite increases in physical memory and heap
size. When garbage collection does occur, it concerns every thread.  Since
garbage collection does not happen often (examples are given in
{\color{red} Chapter 9}), the design philosophy is to keep garbage
collection as simple as possible under concurrent execution.  To allow safe
access to data in the heaps, garbage collection suspends all running
threads except the thread which triggered it, referred to as the GC
thread. The GC thread performs a conventional garbage collection, during
which the program runs sequentially.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[font=\small\tt]
  %\draw[help lines] (0,0) grid (16,8);
  \def\vsp{0.5cm}
  %
  % draw the program text
  \begin{scope}[anchor=west]
    \draw
    (0,7)        node[name=L1]   {MUTEX\_LOCKID(MTX\_NARTHREADS); }
    ++(0,-\vsp)   node[name=L2]   {NARthreads--; }
    ++(0,-\vsp)   node[name=L3]   {MUTEX\_UNLOCKID(MTX\_NARTHREADS); }
    ++(0,2*-\vsp)   node {}
    ++(0,-\vsp)   node[name=L4]   {THREAD\_JOIN(thread); }
    ++(0,2*-\vsp)   node {}
    ++(0,-\vsp)   node[name=L5]   {MUTEX\_LOCKID(MTX\_THREADCONTROL); }
    ++(0,-\vsp)   node[name=L6]   {MUTEX\_LOCKID(MTX\_NARTHREADS); }
    ++(0,-\vsp)   node[name=L7]   {NARthreads++; }
    ++(0,-\vsp)   node[name=L8]   {MUTEX\_UNLOCKID(MTX\_NARTHREADS); }
    ++(0,-\vsp)   node[name=L9]   {MUTEX\_UNLOCKID(MTX\_THREADCONTROL };
  \end{scope}
  %
  % annotate the program
  \begin{scope}[every node/.style={single arrow, draw, thin, align=left,
                                   anchor=west,shape border rotate=180},
                color=black!75!white, font=\it ]
    \draw
    (L2.east) +(5,0)    node[text width=6.3cm]
        {Decrement the counter (Number of Asynchronous Running Threads)}
    (L4.east) +(1.8,0)   node[text width=7.5cm]
        {Do a blocking operation (join in this case),
          a system call (long IO for example) or a delay, etc.}
    (L7.east) +(5,0)    node [text width=6.1cm]
        {Wait for GC to finish, if there is one. Increment the counter.};
  \end{scope}

\end{tikzpicture}
\end{center}
\caption{\label{trackThreads}Tracking the number of running threads in the system}
\end{figure}

\label{GC-Mutex}
The GC thread first locks the garbage collection mutex
\texttt{MTX\_THREADCONTROL}, and then sets a global flag that announces the
need to perform garbage collection. It directs all running threads to
converge to a special routine called \texttt{thread\_control}, the
coordination point for thread suspension and resumption. The global flag is
checked by all threads once per virtual machine instruction.  The runtime
system keeps track of how many threads are running using a global counter
\texttt{NARthreads}, which is incremented and decremented by the threads
depending on their state: before a thread blocks on a mutex it decrements
the counter, and increments it after unblocking. The counter itself is
protected by a mutex and can be incremented only if there is no garbage
collection request pending or taking place (Figure \ref{trackThreads}).

After a call for garbage collection causes a thread to enter the
\texttt{thread\_control} function, the thread decrements the running
threads counter \texttt{NARthreads} and goes to sleep on the \texttt{gc}
condition variable. The GC thread keeps testing the running threads counter
until its value decreases to 1, meaning that the GC thread is the only
thread running. Then the GC thread proceeds to perform the garbage
collection. After finishing, the GC thread unlocks the garbage collection
mutex, and does a broadcast to the \texttt{gc} condition variable. This
wakes up all of the threads blocked on the \texttt{gc} variable, allowing
them to resume their execution and return to where they called the
\texttt{thread\_control} function in the first place.

In garbage collection-intensive applications, it may happen that two (or
more) threads trigger a garbage collection at the same moment. This
situation can be handled in various ways.  The simplest solution is to
block all of the threads requesting garbage collection after the first one
and let the first one proceed and finish as described above (block all of
the other threads and then wake them up again). A better solution, the one
adopted in this design, is to make the threads that are competing for
garbage collection aware of each other. Instead of a sequence of block all
others followed by a wake up others performed by each GC thread, a GC
thread can hand the control over to the next GC thread in line, if there is
one, and then go to sleep. Only the last GC thread in the line wakes up all
of the blocked threads. This strategy eliminates intermediate block/wakeup
operations, leaving only one block others operation by the first GC thread
and one wake up others operation by the last GC thread. A counting
semaphore (gc semaphore) controls the queuing and synchronizing of several
threads to garbage collect.  The first thread to request a garbage
collection does not block on the semaphore, but subsequent requesters
do. After finishing garbage collection, each thread signals the semaphore
to wake up the next GC thread and goes to sleep. The last GC thread to wake
up is responsible for waking up all of the threads after finishing.

An additional measure increases the effectiveness of the strategy just
described: triggering an artificial garbage collection if another thread
has a garbage collection request pending. After a thread triggers a garbage
collection and starts the protocol to suspend all other threads, each
thread answers the call, but before going to sleep on the gc condition
variable it checks if its heap is nearly full. The nearly full value
depends on the heap size, the number of threads, and the nature of the
application. Tests indicate that generally, a value in the range 90\%-95\%
is reasonable for most applications (garbage collection example statistics
are given in {\color{red} Table 9.1} ). Thus if the threads heap has only
5\%-10\% of its total size free, the thread queues up to do garbage
collection instead of going to sleep on the \texttt{gc} condition variable.
This technique forces several garbage collections that might be separated
by very short periods of time to cluster together, requiring only a single
suspension for all of the threads to do the several garbage
collections. Figure \ref{GC-Overview} shows an abstract overview of garbage
collection and concurrency control.

\begin{figure}[h]
\begin{center}
\begin{tikzpicture}[draw=black,very thick,font=\small\tt,>=Latex,
    Thread/.style={text=black,draw=skyblue,
        shading=radial, inner color=blue!40!white, outer color=orange!50!white,
        shape=rectangle, minimum width=2.5cm, minimum height=8mm, rounded corners=2mm},
    % Queue Release Circle and resumption action
    QRC/.style={draw=#1,shape=circle,radius=2.5mm,anchor=north},
    QRes/.style={draw=#1,shape=arrow box,arrow box arrows={south:5mm},
                 arrow box shaft width=4mm,
                 anchor=north,align=center},
    % Thread Queue
    TQ/.style={draw=#1,shape=rectangle, rounded corners=4mm,
                 minimum width=3cm, minimum height=2cm},
    % Action
    Act/.style={draw=#1,shape=rectangle, rounded corners=1.5mm,anchor=north,
                 minimum width=3cm, minimum height=0.7cm,align=center},
    % Branch
    Br/.style={draw=#1,shape=diamond,align=center,inner sep=0,
                 minimum size=2cm,anchor=north}
  ]
  \definecolor{skyblue}{rgb}{0.53, 0.81, 0.92}
  \definecolor{teagreen}{rgb}{0.82, 0.94, 0.75}
  %\draw[help lines] (0,0) grid (16,20);

  % draw the GC queue
  \draw (2,9) node[TQ=black,name=GCQ,anchor=north, text depth=3.75cm] {GC Queue};
  \foreach \n/\y in {1/1.5,2/3.5}
           {\node[Thread,name=gqt\n] at ($(GCQ.north)+(0,-\y)$) {Thread};}
  \draw[dotted] (gqt1) -- (gqt2);
  \draw (GCQ.south)     node[QRC=black,name=GCQ-Out]{};
  \draw (GCQ-Out.south) node[QRes=black,name=GCQ-Res] {Resume one\\thread};
  \draw (GCQ-Res.south east)+(-0.5,-1)
        node[Act=black,name=GCQ-Act, anchor=west] {NARthreads++};
  \draw[->] (GCQ-Res.south) |- (GCQ-Act.west);

  % draw the suspended threads queue
  \draw (14,9) node[TQ=black,name=SusQ,anchor=north, text depth=5.25cm] {Suspended};
  \foreach \n/\y in {1/1.5,2/3,3/5}
           {\node[Thread,name=gqt\n] at ($(SusQ.north)+(0,-\y)$) {Thread};}
  \draw[dotted] (gqt2) -- (gqt3);
  \draw (SusQ.south)     node[QRC=black,name=SusQ-Out]{};
  \draw (SusQ-Out.south) node[QRes=black,name=SusQ-Res] {Resume all\\threads};
  \draw (SusQ-Res.south) node[Act=black,name=SusQ-Act]  {NARthreads++};
  \draw (SusQ-Act)+(0,-1) node[QRC=black,name=SusQ-C1]{};
  \draw[->] (SusQ-Act) -- (SusQ-C1);

  %draw the GC actions
  \draw (7.5,13) node[Act=black,name=Setf] {Set the thread\\call flag};
  \draw (Setf.south)+(0,-0.75) node[Br=black,name=NARq,aspect=4] {NARthreads = 1?};
  \draw (NARq.south)+(0,-0.5) node[Act=black,name=GC] {Perform GC};
  \draw (GC.south)+(0,-0.5)   node[Br=black,name=Emptyq,aspect=2] {GC queue\\empty?};
  \draw[->] (Setf) -- (NARq);
  \draw[->] (NARq) -- (GC);
  \draw[->] (GC)   -- (Emptyq);
  \draw[->] (Emptyq)  |- (SusQ-C1);
  \draw[->] (GCQ-Act) |- (GC);
  \draw[->] (NARq) -- +(3,0) -- +(3,1.5) -- +(0,1.5);

  %draw the GC request actions
  \draw (2,15.75) node[ellipse,draw=red,fill=black!30!white,name=GCReq] {GC request};
  \draw (GCReq.south)+(0,-0.5) node[Br=black,name=Firstq] {first?};
  \draw (Firstq.south)+(0,-1)  node[Act=black,name=GCQ-in] {NARthreads $- -$};
  \draw[->] (GCReq) -- (Firstq);
  \draw[->] (Firstq) -- (GCQ-in);
  \draw[->] (GCQ-in) -- (GCQ);
  \draw[->] (Firstq) -| (Setf);

  %draw the thread call actions
  \draw (14,16) node[Br=black,name=Callq,aspect=2] {Thread\\call?};
  \draw (Callq.south)+(0,-0.75) node[Br=black,name=Fullq,aspect=2] {Near full\\heap?};
  \draw (Fullq.south)+(0,-0.75) node[Act=black,name=SusQ-in] {NARthreads $- -$};
  \draw[->] (Callq) -- (Fullq);
  \draw[->] (Fullq) -- (SusQ-in);
  \draw[->] (SusQ-in) -- (SusQ);

  \draw[->] (Emptyq) -- +(4,0) |- (SusQ-in);
  \draw[->] (Fullq) -- +(-3.5,0) |- (GCReq);

  %draw the running threads
  \draw (8,19) node[rectangle,rounded corners=6mm,fill=teagreen!70!white,
                    minimum height=3cm,minimum width=16cm,name=Ground] {};
  \foreach \n/\x in {1/2, 2/6, n/14}
           { \path (Ground.south west)+(\x,1) node[Thread,name=RTh\n] {Thread \n};}
  \path (Ground.south west)+(10,2) node[Thread,name=RThk] {Thread k};
  \draw (Ground.north west)+(0.25,-0.5) node[font=\Large\tt,anchor=north west]
            {Running threads \dots};
  \draw[dotted] (RTh2) -- (RThn);

  %draw some extra connectors, now all the nodes are defined
  \draw[->] (RThk) -- +(0,-2.5) -| (GCReq);
  \draw[->] (RThn) -- (Callq);
  \draw[->] (SusQ-C1) -| (Ground.south east) +(0,0.4);
  \draw[->] (Callq.west) -- ++(-1,0) -- ++(0,2.9);

  %label the branch decisions
  \draw (Callq.west) node[anchor=south east]{No};
  \draw (Callq.south) node[anchor=north west]{Yes};
  \draw (Fullq.west) node[anchor=south east]{Yes};
  \draw (Fullq.south) node[anchor=north west]{No};
  \draw (Firstq.east) node[anchor=south west]{Yes};
  \draw (Firstq.south) node[anchor=north west]{No};
  \draw (Emptyq.east) node[anchor=south west]{No};
  \draw (Emptyq.south) node[anchor=north west]{Yes};
  \draw (NARq.east) node[anchor=south west]{No};
  \draw (NARq.south) node[anchor=north west]{Yes};

  %draw the signalling paths
  \begin{scope}[<->,dashed,arrows={Triangle[open,reversed,length=3mm,width=4mm]-Latex},auto]
    % could use the intersection library here but, since it's the only case (so far)
    % in the entire book, just use let to pick out the x,y coordinates.
    \draw let \p1 = (SusQ-Out) in let \p2 = (Emptyq) in
       (\x2,\y1) to node{"Wake up all"} node[swap]{signal} (SusQ-Out);
    \draw ($(Emptyq)+(2.5,0)$) |- node[near end]{signal}
                                node[swap,near end]{"Wake up one"} (GCQ-Out);
  \end{scope}

\end{tikzpicture}
\end{center}
\caption{\label{GC-Overview}
  A high level view of the dynamics of suspending/resuming threads for GC}
\end{figure}
\PrimaryIndexEnd{Garbage collection}
