\begin{thebibliography}{ABCD}
  % In book.cls, thebibliography environment is a chapter*{\bibname}
  \addcontentsline{toc}{chapter}{Bibliography}

  \bibitem[Alg12]{Alg12} Jafar Al Gharaibeh.  \emph{Programming Language Support
    for Virtual Environments}. Ph.D. Dissertation, University of Idaho, 2012.

  \bibitem[And83]{And83} Gregory Andrews and Fred Schneider.
    \emph{Concepts and Notations for Concurrent Programming}.
    ACM Computing Surveys. 15:1, March 1983, pp 3-43.

  \bibitem[Berk82]{Berk82} Berk, T., Brownstein, L., and Kaufman, A.
    \emph{A New Color-Naming System for Graphics Languages}.
    IEEE Computer Graphics \& Applications, pp 37-44, May 1982.

  \bibitem[Boehm88]{Boehm88} Barry W. Boehm.
    \emph{A Spiral Model of Software Development and Enhancement}.
    IEEE Computer, vol. 21 no. 5, pp 61-72. 1988.

  \bibitem[But97]{But97} David R. Butenhof.
    \emph{Programming with Posix Threads}. Addison Wesley, 1997.

\bibitem[Clark85]{Clark85} D.~D. Clark.
The {S}tructuring of {S}ystems {U}sing {U}pcalls.                     
In {\em Proceedings of the Tenth ACM Symposium on Operating System    
  Principles}, pages 171--180, Dec. 1985.

  \bibitem[Elm89]{Elm89} Ramez Elmasri and Shamkant B. Navathe.
    \emph{Fundamentals of Database Systems}.
    The Benjamin/Cummings Publishing Company, Redwood City, CA. 1989.

  \bibitem[Eri98]{Eri98} Hans-Erik Eriksson and Magnus Penker.
    \emph{UML Toolkit}. Wiley, New York, NY.  1998.

  \bibitem[Fol95]{Fol95} James D. Foley, Andries Van Dam, Steven K. Feiner, and John
    F. Hughes.  \emph{Fundamentals of Interactive Computer Graphics: Principles
      and Practice in C}. Reading, MA: Addison-Wesley Publishing Company, 1995.

  \bibitem[Gai39]{Gai39} Helen Fouche Gaines.
    \emph{Cryptanalysis: a Study of Ciphers and Their Solution}. Dover 1939.

\bibitem[Gris86]{Gris86} R.~E. Griswold and M.~T. Griswold.
{\em The Implementation of the Icon Programming Language}.            
Princeton University Press, Princeton, New Jersey, 1986.              

  \bibitem[Gris96]{Gris96} Ralph E. Griswold and Madge T. Griswold.
    \emph{The Icon Programming Language, 3rd ed}. Peer-to-Peer Communications,
    San Jose, CA. 1996.

  \bibitem[GJT98]{GJT98} Ralph E. Griswold, Clinton L. Jeffery, and Gregg
    M. Townsend. \emph{Graphics Programming in Icon}. Peer-to-Peer
    Communications, San Jose, CA. 1998.

  \bibitem[GPP71]{GPP71} Griswold, Poage, and Polonsky.
    \emph{The SNOBOL 4 Programming Language, 2nd ed}.
    Englewood Cliffs, N.J. Prentice-Hall, Inc. 1971.

  \bibitem[Jeff99]{Jeff99} Clinton L. Jeffery.
    \emph{Program Monitoring and Visualization: An Exploratory Approach}.
    Springer-Verlag, New York, NY. 1999.

  \bibitem[Lev90]{Lev90} John R. Levine, Tony Mason, and Doug Brown.
    \emph{lex \& yacc}. O'Reilly and Associates, Sebastopol, CA, 1990.

\bibitem[Ogle90]{Ogle90} D.~M. Ogle, K.~Schwan, and R.~Snodgrass.
\emph{The {D}ynamic {M}onitoring of {D}istributed and {P}arallel {S}ystems}.
Technical Report GIT-ICS-90/23, School of Information and Computer    
  Science, Georgia Institute of Technology, Dec. 1990.

% \bibitem[]{} OpenGL Architecture Review Board; Woo, Mason; Neider, Jackie; Davis,
  \bibitem[Woo99]{Woo99} Woo, Mason; Neider, Jackie; Davis, Tom; Shreiner, Dave.
    \emph{OpenGL Programming Guide: the Official Guide to Learning OpenGL, Third
      Edition}. Addison-Wesley, 1999.

% \bibitem[]{} OpenGL Architecture Review Board; Shreiner, Dave. \emph{OpenGL Programming
  \bibitem[Shr00]{Shr00} Shreiner, Dave.
   \emph{OpenGL Programming Guide: the Official
    Reference Document to OpenGL, Third Edition}.  Addison-Wesley, 2000.

  \bibitem[Pax95]{Pax95} Vern Paxson. {Flex: A fast scanner generator, ed. 2.5},
    \url{www.gnu.org/manual/flex/}, 1995.

  \bibitem[Ram94]{Ram94} Norman Ramsey. \emph{Literate programming simplified}.
    IEEE Software, 11(5):97-105, September 1994.

\bibitem[Town89]{Town89} R.~E. Griswold and G.~M. Townsend.
\emph{The {V}isualization of {D}ynamic {M}emory {M}anagement in the {I}con  
  {P}rogramming {L}anguage}.
\newblock Technical Report 89-30, Department of Computer Science, University of 
  Arizona, Dec. 1989.

  \bibitem[Wall91]{Wall91} Larry Wall and Randall Schwartz. \emph{Programming
    \index{Perl}Perl}.  O'Reilly and Associates, Sebastopol, CA, 1991.

\end{thebibliography}
